Von Neumann-Morgenstern and Savage Theorems for Causal Decision Making

### 2020-08-30 Theory_and_decision/Draft/Theory_and_decision_draft.tex
cambios en limitations of theorems

\subsection{Limitations of the Theorems}
\label{theorems_limitations}
The mentioned theorems have the limitations of relying only on associative information; this is, on correlations and patterns found in data rather than strong causal relations. For example, as mentioned by \cite{pearl2009causality}, using only associative information one would be tempted to stop going to the doctor in order to stop being diagnosed as sick. On the other hand, our results provide an easy to implement way of making choices using causal relations and beliefs about such relations as used by humans. Human beings \textit{know} (this is, use causal relations) that going to the physician does not \textit{cause} being sick and this is what Pearl calls \textit{commonsensical} decision making. Also, as we will mention later in the paper, associative information lacks an \textit{ordering} thus failing to determine \textit{if $X$ is performed, then $Y$ will be caused}.

Formally, in this paper we consider the actions chosen by a decision maker as an \textit{interventions} over a causal graphical model; this is, in the classical von Neumann-Morgenstern and Savage Theorems, once a decision maker chooses an action $a$, the observed consequence has a probability $P(c|a)$; if causal information is considered, then the probability of a certain consequence is $P(c|do(a))$. In the next section we state some basic facts on causality and why $P(\cdot |x)$ is different from $P(\cdot | do(x))$ for some value $x$.

One of the fundamental aspects of Causality is the presence of an order and a context; the latter of such elements not being present in associative information; this is, consider a distribution $p(a,t)$, which we can factorize in the two following ways:
\begin{equation}
    p(a,t)=p(a|t)p(t)
\end{equation}
and
\begin{equation}
    p(a,t)=p(t|a)p(a)
\end{equation}
which can be represented indistinguishably as Bayesian Networks as $a \to t$ or as $t \to a$. This situation does not happen while considering causal relations.

Also, we recall the difference between an observation and an intervention: $p(x|y) \neq p(x | do(y))$. If this was not the case, then correlation would equal causation which can be shown in a variety of examples to be false, see \cite{pearl2009causality}.

\section{Causal Decision Problems}
\label{causal_decision_problems}
\subsection{On causality}
\label{on_causality}
The concept of Causality deals with regularities found in a given environment which are stronger than probabilistic (or associative) relations in the sense that a causal relation allows for evaluating a change in the \textit{consequence} given that a change in the \textit{cause} is performed. The \textit{manipulationist} interpretation of Causality (details in \cite{woodward2005making}) is adopted here. The main paradigm is clearly expressed by \cite{campbell1979quasi} as \textit{manipulation of a cause will result in a manipulation of the effect}. Consider the following example from \cite{woodward2005making}: manually forcing a barometer to go down won't cause a storm, whereas the occurrence of a storm will cause the barometer to go down.

We know from \cite{granger1969investigating} and \cite{lamport1978time} that there are two sufficient and necessary conditions for causality: a context and an order. We will adopt here the formal definition of Causality given by \cite{spirtes2000causation}; i.e., a stochastic relation between events which is \textit{irreflexive, antisymmetric} and \textit{transitive}. Similar descriptions of the manipulationist approach can be found in \cite{holland1986statistics}. Causal inference tools, such as Pearl's do-calculus \citep{pearl2009causality} allow to find the effect of an intervention in terms of probabilistic information when certain conditions are met. For what remains, we assume the \textit{causal axioms} found in \citep{spirtes2000causation} as well as the condition known as \textit{causal sufficiency} \citep{pearl2009causality}.

>>>>

\subsection{Limitations of the Theorems}
\label{theorems_limitations}
The mentioned theorems have the limitations of relying only on associative information; this is, on correlations and patterns found in data rather than strong causal relations. For example, as mentioned by \cite{pearl2009causality}, using only associative information one would be tempted to stop going to the doctor in order to stop being diagnosed as sick. On the other hand, our results provide an easy to implement way of making choices using causal relations and beliefs about such relations as used by humans. Human beings \textit{know} (this is, use causal relations) that going to the physician does not \textit{cause} being sick and this is what Pearl calls \textit{commonsensical} decision making.

We know from \cite{granger1969investigating} and \cite{lamport1978time} that there are two sufficient and necessary conditions for causality: a context and an order.; the latter of such elements not being present in associative information; this is, consider a distribution $p(a,t)$, which we can factorize in the two following ways:
\begin{equation}
    p(a,t)=p(a|t)p(t)
\end{equation}
and
\begin{equation}
    p(a,t)=p(t|a)p(a)
\end{equation}
which can be represented indistinguishably as Bayesian Networks as $a \to t$ or as $t \to a$. This situation does not happen while considering causal relations. Also, associative information lacks an \textit{ordering} thus failing to determine \textit{if $X$ is performed, then $Y$ will be caused}. We recall the difference between an observation and an intervention: $p(x|y) \neq p(x | do(y))$. If this was not the case, then correlation would equal causation which can be shown in a variety of examples to be false, see \cite{pearl2009causality}.

Formally, in this paper we consider the actions chosen by a decision maker as an \textit{interventions} over a causal graphical model; on the other hand, in the classical von Neumann-Morgenstern and Savage Theorems, once a decision maker chooses an action $a$, the observed consequence has a probability $P(c|a)$; if causal information is considered, then the probability of a certain consequence is $P(c|do(a))$. In the next section we state some basic facts on causality and why $P(\cdot |x)$ is different from $P(\cdot | do(x))$ for some value $x$.

\section{Causal Decision Problems}
\label{causal_decision_problems}
\subsection{On causality}
\label{on_causality}
The concept of Causality deals with regularities found in a given environment (context) which are stronger than probabilistic (or associative) relations in the sense that a causal relation allows for evaluating a change in the \textit{consequence} given a change in the \textit{cause}. We adopt here the \textit{manipulationist} interpretation of Causality (details in \cite{woodward2005making}). The main paradigm is clearly expressed by \cite{campbell1979quasi} as \textit{manipulation of a cause will result in a manipulation of the effect}. Consider the following example from \cite{woodward2005making}: manually forcing a barometer to go down won't cause a storm, whereas the occurrence of a storm will cause the barometer to go down.

We will adopt here the formal definition of Causality given by \cite{spirtes2000causation}; i.e., a stochastic relation between events which is \textit{irreflexive, antisymmetric} and \textit{transitive}. Similar descriptions of the manipulationist approach can be found in \cite{holland1986statistics}. Causal inference tools, such as Pearl's do-calculus \citep{pearl2009causality} allow to find the effect of an intervention in terms of probabilistic information when certain conditions are met. For what remains, we assume the \textit{causal axioms} found in \citep{spirtes2000causation} as well as the condition known as \textit{causal sufficiency} \citep{pearl2009causality}.
